---
title: "Storms"
subtitle: "A NOAA database analysis for Health and Economic consequences"
author: "Mathieu C."
date: "2 octobre 2019"
output: html_document
---
##Synopsis
The file "Storm Data" comes from the U.S. National Oceanic and Atmospheric  
Administration's (NOAA) and is accessible at [this url](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) on oct. 2nd 2019.  


## Data Processing

  
We will begin by taking a look at the data:
```{r cache=TRUE}
# Downloading file (47Mb, might take a while depending on the connection and speeds)
# note that this step is cached and will be executed much faster if repeated:
myurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if (!file.exists("Storm_Data.csv.bz2")){
        download.file(myurl, destfile = "Storm_Data.csv.bz2")
}
# Reading in the dataframe, again, might take a while.
if (!exists("StormData")){
        StormData <- read.csv("Storm_Data.csv.bz2")
}
str(StormData)
```
From the above call to `str()` function, the database seems pretty organized.  
Let's check for missing values:
```{r}
na_check <- lapply(StormData, is.na)
na_check <- lapply(na_check, sum)
str(na_check)
```
We can see that the database is an "all or nothing" type: either a variable  
has (almost) all it's values filled in, or almost all the data is missing.  
  
Fortunately, in the "missing" camp, we have only two variables `F` and `COUNTYENDN`. These variables don't seem to be relevant to the questions at hands.


## Data Analysis

```{r}
library(dplyr)
library(ggplot2)
library(reshape2)
df_by_event <- StormData %>%
        group_by(EVTYPE) %>%
        summarize(fatalities = sum(FATALITIES, na.rm = T),
                  injuries = sum(INJURIES, na.rm = T),
                  total.inj.fat = fatalities + injuries)
df_fatalities <- head(df_by_event[order(df_by_event$fatalities,decreasing = T),],20)
df_injuries <- head(df_by_event[order(df_by_event$injuries,decreasing = T),],20)
df_cumul_health <- head(df_by_event[order(df_by_event$total.inj.fat,decreasing = T),],20)
df_cumul_health <- transform(df_cumul_health, EVTYPE = tolower(as.character(EVTYPE)))
plotgrp <- melt(df_cumul_health[,-4], id.vars = 'EVTYPE')
plotgrp <- transform(plotgrp, EVTYPE = factor(EVTYPE, levels=unique(EVTYPE)))

g1 <- ggplot(plotgrp, aes(x=EVTYPE, y = log(value,2), fill = variable))
g1 + geom_bar(stat = "identity", position = "dodge") +
        theme(axis.text.x = element_text(angle = 90, 
                                         vjust = 0.25,
                                         hjust = 0.95)) +
        geom_text(aes(label=value), 
                  position=position_dodge(width=0.9), 
                  hjust = 1, 
                  angle = 90)+
        ylab("Log 2 scale of events")+
        xlab("Event")+
        labs(title = paste0("Top 20 events in decreasing order of cumulative",
                            " injuries & fatalities;\n Sum of reported events",
                            " in years 1950 to 2011"),
             subtitle = "Warning: Scale is Logarithmic but absolute values are displayed on the bars")
        
```

